---
title: "Data Cleaning and Preparation for MPS1 Inhibitor Modeling"
author: "Hoa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


## Introduction 
The objective of this project is to develop a machine learning-based classification model to identify potential inhibitors of the TTK (MPS1) kinase using publicly available bioactivity data from the ChEMBL database.

## 1. Load Libraries
```{r}
library(dplyr)
library(ggplot2)

```

## 2. Data Loading
```{r}
#Load and inspect dataset
data <- read.csv("CHEMBL_bioactivity_mps1.csv")

# Quick inspection
#head(data)
dim(data)
```

## 3. Data Cleaning 


### 3.1. Filter IC50 values
```{r}
# Keep only IC50 in nM and exact relation
data_ic50 <- data %>%
  filter(Standard.Type == "IC50",
         Standard.Units == "nM",
         Standard.Relation == "'='") %>%
  filter(!is.na(Standard.Value)) %>%
  distinct(Smiles, .keep_all = TRUE)
```
Only IC50 values reported in nM with exact relation were retained. Duplicate compounds were removed based on their SMILES representation to avoid data leakage.


```{r}
# Convert numeric variables and remove NA rows 
data_ic50 <- data_ic50 %>%
  mutate(
    Standard.Value = as.numeric(Standard.Value),
    Molecular.Weight = as.numeric(Molecular.Weight),
    AlogP = as.numeric(AlogP)
  ) %>%
  filter(
    !is.na(Standard.Value),
    !is.na(Molecular.Weight),
    !is.na(AlogP)
  )
```

```{r}
dim(data_ic50)
cat("Number of compounds after filtering:", nrow(data_ic50))
```


### 3.2. Check Distributions
```{r}
#Molecular Weight
p_mw <- ggplot(data_ic50, aes(x = Molecular.Weight)) +
  geom_histogram(bins = 30,
                 fill = "#4C78A8",
                 color = "black",
                 alpha = 0.85) +
  labs(title = "Molecular Weight Distribution",
       x = "Molecular Weight (Da)",
       y = "Count") +
  theme_classic(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_mw

#AlogP
p_logp <- ggplot(data_ic50, aes(x = AlogP)) +
  geom_histogram(bins = 30,
                 fill = "#E45756",
                 color = "black",
                 alpha = 0.85) +
  labs(title = "AlogP Distribution",
       x = "AlogP",
       y = "Count") +
  theme_classic(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_logp


#Save plots
ggsave("mw_distribution.png", p_mw,
       width = 5.5, height = 3.5, dpi = 300, bg = "white")

ggsave("alogp_distribution.png", p_logp,
       width = 5.5, height = 3.5, dpi = 300, bg = "white")

```
The distributions appear pretty normal, with no extreme outliers observed; therefore, no trimming was necessary.

### 3.3. Create Activity Label
```{r}
# Compounds with IC50 ≤ 1000 nM classified as active, while compounds with IC50 > 1000 nM classified as inactive.
data_ic50$Activity <- ifelse(data_ic50$Standard.Value <= 1000, 1, 0 )
dim(data_ic50)
```
The binary label will serve as the target variable for modeling.

### 3.4. Select relevant columns for modeling
```{r}
# Molecule.ChEMBL.ID, Smiles, Activity, Molecular.Weight, AlogP, and #RO5.Violations
data_model <- data_ic50 %>%
  select(
    Molecule.ChEMBL.ID,
    Smiles,
    Molecular.Weight,
    AlogP,
    X.RO5.Violations,
    Activity
  )

# Dataset summary after cleaning
#head(data_model)
dim(data_model)
#table(data_model$Activity)
summary(data_model)
```


```{r}
cat("Number of compounds after cleaning:", nrow(data_model))
prop.table(table(data_model$Activity))
```

### 3.5. Activity Class Distribution 
```{r}
# Count activity classes

activity_counts <- data_model %>%
  count(Activity) %>%
  mutate(Activity = factor(Activity,
                           levels = c(0,1),
                           labels = c("Inactive", "Active")))

p <- ggplot(activity_counts, aes(x = Activity, y = n, fill = Activity)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(title = "Activity Distribution",
       x = "Class",
       y = "Number of Compounds") +
  scale_fill_manual(values = c("#4C78A8", "#E45756")) +
  theme_classic(base_size = 14) +   # removes grid
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black")
  )

p

#Save graph
ggsave("activity_distribution.png",
       plot = p,
       width = 6,
       height = 4,
       dpi = 300,
       bg = "white") 
```
The class distribution is highly imbalanced, which will be addressed during model training.

### 3.6. Save Clean Dataset
```{r}
# Save dataset
write.csv(data_model, "bioactivity_mps1_clean.csv", row.names = FALSE)

```
